---
pagetitle: "breast_cancer_ml"
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float:
     collapsed: false
     smooth_scroll: false
---

<style>
  body {
    font-family: Georgia, "Times New Roman", Times, serif;
  }

  h1, h2 {
    color: #333;
  }
</style>

<link rel="stylesheet" href="styles.css" type="text/css">
<link rel="stylesheet" href="site_libs/academicons-1.9.1/css/academicons.min.css"/>

<br><br><br>

## **Breast Cancer Prediction Analysis using ML Methods**

![](https://img.shields.io/badge/Using-R-blue)
![](https://img.shields.io/badge/Method-Logistic%20%7C%20LASSO%20%7C%20Tree%20%7C%20RF%20%7C%20XGBoost-orange)
![](https://img.shields.io/badge/Data-UCI_Breast_Cancer_Wisconsin-green)
![](https://img.shields.io/badge/Evaluation-ROC%2FAUC%20%7C%20Sensitivity%20%7C%20Confusion_Matrix-purple)


<br>


<br><br><br>

### 1. Introduction
<br>
Breast cancer is one of the most prevalent cancers worldwide, making early and accurate diagnosis critical for improving patient outcomes. <br>

This project applies a comparative machine learning approach to the UCI Breast Cancer Wisconsin dataset, evaluating multiple classification models to distinguish malignant from benign tumors.

<br><br><br>

### 2. Goal
<br>
This project develops and compares five classification models, including Logistic Regression, LASSO, Decision Tree, Random Forest, and XGBoost, for breast cancer prediction. <br>

Model performance is evaluated with a focus on sensitivity and ROC/AUC, given the clinical importance of minimizing false negatives.


<br><br><br>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(mlbench)
library(caret)
library(dplyr)
library(ggplot2)
library(tidyr)
library(corrplot)
library(glmnet)
library(randomForest)
library(rpart)
library(rpart.plot)
library(xgboost)
library(pROC)
library(car)
data(BreastCancer)
```


### 3. Exploratory Data Analysis
<br><br>

#### 3-1. Data Preprocessing

<br><br>
The raw dataset contains **699 observations and 11 variables**, including one patient ID variable, 9 cytological predictors scored on a 1–10 scale, and one binary target variable (`Class`).
<br><br><br>
Three preprocessing steps were applied prior to analysis:

<br>
**1. Missing value removal**  
16 observations with missing values in `Bare.nuclei` were removed using `na.omit()`, 
resulting in a final dataset of **683 observations**.
<br><br>

**2. ID column removal**  
The `Id` variable was excluded as it serves as a patient identifier 
and carries no predictive information.
<br><br>

**3. Type conversion**  
All predictors were originally stored as ordered factors with string levels (`"1"` to `"10"`). To enable numerical computation, they were converted to numeric using `as.numeric(as.character(.))`. The intermediate `as.character()` step prevents R from returning factor level indices instead of the actual values.

<br><br>

```{r preprocessing}
dim(BreastCancer)

dataset <- BreastCancer %>%
  na.omit() %>%
  select(-Id) %>%
  mutate(across(-Class, ~ as.numeric(as.character(.))))

dim(dataset)
anyNA(dataset)
glimpse(dataset)
```

<br><br><br><br>

#### 3-2. Class Distribution

<br>
The target variable `Class` is binary, consisting of **benign** and **malignant** tumor diagnoses. <br>
As shown in the plot below, the dataset contains **444 benign (65.0%)** and **239 malignant (35.0%)** 
observations, indicating a moderate class imbalance.

To account for this imbalance, stratified sampling will be applied during the train/test split 
to ensure both subsets preserve the original class ratio. Given this class imbalance, sensitivity and ROC/AUC were prioritized over overall accuracy as the primary evaluation metrics.
<br><br><br>

```{r class_distribution}
ggplot(dataset, aes(x = Class, fill = Class)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5) +
  theme_minimal() +
  labs(title = "Class Distribution", x = "Diagnosis", y = "Count") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  guides(fill = "none")
```


<br><br><br><br>


#### 3-3. Feature Distributions

<br>
To examine how each predictor differs across tumor classes, density plots and boxplots were generated for all nine features.

Density plots show the overall shape of each feature distribution and the degree of overlap between benign and malignant cases. Boxplots complement this by highlighting differences in medians, interquartile ranges, and potential outliers.

<br><br><br>


```{r feature_distributions, fig.width=10, fig.height=7}
dataset %>%
  select(where(is.numeric), Class) %>%
  pivot_longer(-Class, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value, fill = Class)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal() +
  labs(title = "Feature Distributions by Class")
```

<br><br>

```{r boxplot, fig.width=10, fig.height=7}
dataset %>%
  select(where(is.numeric), Class) %>%
  pivot_longer(-Class, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = Class, y = value, fill = Class)) +
  geom_boxplot(outlier.shape = 21, outlier.size = 1.5) +
  facet_wrap(~ variable, scales = "free_y") +
  theme_classic() +
  labs(title = "Feature Distributions by Class (Boxplot)") +
  guides(fill = "none")
```

<br>

Based on the plots, features such as `Bare.nuclei`, `Cell.shape`, `Cell.size`, and `Bl.cromatin` appear to show relatively clear separation between benign and malignant cases, suggesting that they may serve as strong predictors in subsequent models. In contrast, `Mitoses` showed substantial overlap between classes, indicating weaker discriminative power.


<br><br><br><br>

#### 3-4. Correlation Analysis

<br>
A correlation matrix was computed to examine pairwise relationships among the nine predictors and to assess potential multicollinearity prior to modeling.


<br><br>

```{r correlation_matrix, fig.width=8, fig.height=7}
cor_matrix <- dataset %>%
  select(where(is.numeric)) %>%
  cor()

corrplot(cor_matrix,
         method      = "color",
         type        = "upper",
         addCoef.col = "black",
         number.cex  = 0.7,
         tl.cex      = 0.8,
         diag        = FALSE,
         title       = "Correlation Matrix of Predictors",
         mar         = c(0,0,1,0))
```

<br>

The strongest correlation was observed between `Cell.size` and `Cell.shape` (r = 0.91), suggesting that these variables capture highly similar information. In addition, `Cell.size` showed strong correlations with several other predictors, indicating it may contribute to multicollinearity.

In contrast, `Mitoses` exhibited consistently weak correlations with other features, suggesting limited redundancy.

These findings motivated a formal assessment of multicollinearity using Variance Inflation Factors (VIF) in the following section.






<br><br><br><br>


### 4. Multicollinearity Check (VIF)

<br>

Although the correlation matrix indicated several moderately high pairwise correlations,
a formal multicollinearity assessment was conducted using the **Variance Inflation Factor (VIF)**. 
Unlike pairwise correlations, VIF measures how much the variance of each coefficient is inflated due to its relationships with other predictors.

<br>

A logistic regression model was fitted on the full dataset solely for the purpose of VIF computation, 
prior to any train/test split. VIF values were interpreted using the following thresholds:

- VIF < 5 : acceptable
- VIF 5–10 : moderate concern
- VIF > 10 : severe multicollinearity

<br>

```{r vif, fig.width=7, fig.height=5}
vif_model <- glm(Class ~ ., data = dataset, family = binomial)

vif_result <- vif(vif_model)  
print(vif_result)
```

<br>

```{r vif_plot, fig.width=7, fig.height=5}

vif_df <- data.frame(
    Variable = names(vif_result),
    VIF      = as.numeric(vif_result)
)

ggplot(vif_df, aes(x = reorder(Variable, VIF), y = VIF, fill = VIF > 5)) +
    geom_col() +
    geom_hline(yintercept = 5,  linetype = "dashed", color = "orange") +
    geom_hline(yintercept = 10, linetype = "dashed", color = "red") +
    coord_flip() +
    scale_fill_manual(values = c("steelblue", "tomato"),
                      labels = c("VIF ≤ 5", "VIF > 5"),
                      name   = "") +
    theme_minimal() +
    labs(title = "Variance Inflation Factor (VIF)",
         x = "", y = "VIF")
```



<br>

All nine predictors returned VIF values below 5, indicating no evidence of severe multicollinearity when analyzed collectively. However, the exceptionally high pairwise correlation between `Cell.size` and `Cell.shape` (r = 0.91) still requires careful interpretation.

As will be observed in Section 5-1, `Cell.size` yields a negative and statistically non-significant coefficient within the standard logistic regression framework. This outcome may partly reflect coefficient instability associated with the high inter-correlation between these predictors, even when formal VIF thresholds are not exceeded. 

These findings directly inform the application of LASSO regularization in Section 5-2, which utilizes data-driven shrinkage to enhance the stability and reliability of the parameter estimates while performing autonomous variable selection.

<br><br><br><br>



### 5. Model Development

<br>

All models were trained on 70% of the data and evaluated on the remaining 30%, 
using stratified sampling to preserve the original class ratio across both subsets.

<br>

```{r train_test_split}
set.seed(42)
train_idx <- createDataPartition(dataset$Class, p = 0.7, list = FALSE)
train <- dataset[train_idx, ]
test  <- dataset[-train_idx, ]
```

<br><br><br>

#### 5-1. Logistic Regression

<br>

Logistic regression was used as the baseline model, incorporating all nine predictors. <br>
As a parametric model, it provides interpretable coefficients, allowing direct assessment of the direction and magnitude of each predictor's association with malignancy.

<br>

```{r logistic}
logit_model <- glm(Class ~ ., data = train, family = binomial)
summary(logit_model)
```

<br>

```{r logistic_pred}
logit_prob <- predict(logit_model, newdata = test, type = "response")
logit_pred <- ifelse(logit_prob > 0.5, "malignant", "benign")
logit_pred <- factor(logit_pred, levels = levels(test$Class))
confusionMatrix(logit_pred, test$Class, positive = "malignant")
```

<br>

Among the predictors, `Cl.thickness`, `Marg.adhesion`, `Bare.nuclei`, `Bl.cromatin`, and `Normal.nucleoli` were statistically significant (p < 0.05).

`Cell.size` returned a negative coefficient despite its strong association with malignancy observed in EDA. This is likely due to its high correlation with `Cell.shape` (r = 0.91), which absorbed much of its predictive contribution.

This instability in coefficient estimation motivated the use of LASSO in the following section.

The model achieved an **accuracy of 97.6%** and a **sensitivity of 95.8%**, misclassifying three malignant cases as benign.

<br><br><br>

#### 5-2. LASSO

<br>

LASSO (Least Absolute Shrinkage and Selection Operator) was applied to address the coefficient instability observed in logistic regression. By imposing an L1 penalty, LASSO shrinks less informative coefficients toward zero, effectively performing automatic variable selection.

The optimal penalty parameter lambda was selected via 10-fold cross-validation, using `lambda.1se`, the largest lambda within one standard error of the minimum, which favors a simpler model without substantial loss in predictive performance.

<br>

```{r lasso}
x_train <- model.matrix(Class ~ ., data = train)[, -1]
x_test  <- model.matrix(Class ~ ., data = test)[, -1]
y_train <- ifelse(train$Class == "malignant", 1, 0)
y_test  <- ifelse(test$Class  == "malignant", 1, 0)

set.seed(42)
lasso_cv <- cv.glmnet(x_train, y_train,
                      family  = "binomial",
                      alpha   = 1,
                      nfolds  = 10)
plot(lasso_cv)
```

<br>

```{r lasso_coef}
coef(lasso_cv, s = "lambda.1se")
```

<br>

```{r lasso_pred}
lasso_prob <- predict(lasso_cv, newx = x_test, 
                      s    = "lambda.1se", 
                      type = "response")
lasso_pred <- ifelse(lasso_prob > 0.5, "malignant", "benign")
lasso_pred <- factor(lasso_pred, levels = levels(test$Class))

confusionMatrix(lasso_pred, test$Class, positive = "malignant")
```

<br>

LASSO eliminated `Mitoses` from the model by shrinking its coefficient to zero, consistent with its weak discriminative power observed in EDA.

`Cell.size`, which showed a negative coefficient in logistic regression, was assigned a positive coefficient under LASSO, reflecting its true positive association with malignancy.

The model achieved an **accuracy of 97.1%** and a **sensitivity of 94.4%**, representing a slight decrease compared to logistic regression. However, the resulting model is more parsimonious and provides more stable coefficient estimates, making it preferable when interpretability is a priority.


<br><br><br>


#### 5-3. Decision Tree

<br>

A decision tree was used to provide an interpretable, rule-based classification model. <br>
Unlike logistic regression, decision trees require no assumptions about linearity or variable distributions, and the resulting structure can be directly translated into clinical decision rules.

To prevent overfitting, pruning was applied by selecting the optimal complexity parameter (cp) based on cross-validated error rates.

<br>

```{r decision_tree}
tree_model <- rpart(Class ~ .,
                    data    = train,
                    method  = "class",
                    control = rpart.control(cp = 0.01))

printcp(tree_model)
plotcp(tree_model)
```

<br>

```{r tree_pruned}
best_cp     <- tree_model$cptable[which.min(tree_model$cptable[, "xerror"]), "CP"]
tree_pruned <- prune(tree_model, cp = best_cp)

rpart.plot(tree_pruned,
           type  = 4,
           extra = 104,
           main  = "Pruned Decision Tree")
```

<br>

```{r tree_pred}
tree_pred <- predict(tree_pruned, newdata = test, type = "class")
confusionMatrix(tree_pred, test$Class, positive = "malignant")
```

<br>

The pruned tree utilized only two predictors, `Cell.size` and `Cell.shape`, to construct three terminal nodes. <br>
A simple decision rule emerged: observations with `Cell.size` below 3 were classified as benign with a probability of 97%, while those with both `Cell.size` and `Cell.shape` at or above 3 were classified as malignant with a probability of 92%.

The model achieved an **accuracy of 94.1%** and a **sensitivity of 91.6%**, lower than that of logistic regression and LASSO. <br>
This reduction in performance reflects the trade-off between interpretability and predictive power. 
The tree relies on only two predictors, discarding information from the remaining seven features.


<br><br><br>

#### 5-4. Random Forest

<br>

Random Forest was used as an ensemble method to improve predictive performance beyond a single decision tree.<br> 
It aggregates predictions from 500 decision trees, each trained on a random subset of features, thereby reducing variance and improving generalization.
<br>


```{r random_forest}
set.seed(42)
rf_model <- randomForest(Class ~ .,
                         data       = train,
                         ntree      = 500,
                         mtry       = floor(sqrt(ncol(train) - 1)),
                         importance = TRUE)

varImpPlot(rf_model,
           main = "Random Forest: Variable Importance",
           pch  = 16, col = "darkblue")
```

<br>

```{r rf_pred}
rf_prob <- predict(rf_model, newdata = test, type = "prob")[, "malignant"]
rf_pred <- predict(rf_model, newdata = test, type = "response")

confusionMatrix(rf_pred, test$Class, positive = "malignant")
```

<br>

The variable importance plot shows that `Bare.nuclei`, `Cell.shape`, and `Cell.size` are consistently ranked among the most influential predictors across both MeanDecreaseAccuracy and MeanDecreaseGini metrics, consistent with findings from EDA and the decision tree model.

Unlike the decision tree, Random Forest leverages all nine predictors, allowing it to capture more complex relationships in the data.

The model achieved an **accuracy of 98.0%** and a **sensitivity of 97.2%**, slightly outperforming logistic regression and reducing false negatives from three to two cases, while providing greater robustness to overfitting.

  
<br><br><br>


#### 5-5. XGBoost

<br>

XGBoost (Extreme Gradient Boosting) was used as a gradient boosting method that builds trees sequentially, with each tree correcting the errors of the previous one. 
In contrast to Random Forest, which builds trees independently, XGBoost optimizes a regularized objective function and often achieves higher predictive performance.

Key hyperparameters included a learning rate (eta) of 0.1, a maximum tree depth of 4, and a subsampling rate of 0.8.<br> 
Early stopping was applied based on validation AUC, halting training if no improvement was observed for 10 consecutive rounds.

<br>

```{r xgboost}
dtrain <- xgb.DMatrix(data = x_train, label = y_train)
dtest  <- xgb.DMatrix(data = x_test,  label = y_test)

params <- list(
  objective        = "binary:logistic",
  eval_metric      = "auc",
  eta              = 0.1,
  max_depth        = 4,
  subsample        = 0.8,
  colsample_bytree = 0.8
)

set.seed(42)
xgb_model <- xgb.train(params                = params,
                        data                  = dtrain,
                        nrounds               = 100,
                        evals                 = list(train = dtrain, test = dtest),
                        verbose               = 1,
                        early_stopping_rounds = 10)
```

<br>

```{r xgb_imp}
xgb_imp <- xgb.importance(feature_names = colnames(x_train), model = xgb_model)
xgb.plot.importance(xgb_imp, main = "XGBoost: Variable Importance")
```

<br>

```{r xgb_pred}
xgb_prob <- predict(xgb_model, dtest)
xgb_pred <- ifelse(xgb_prob > 0.5, "malignant", "benign")
xgb_pred <- factor(xgb_pred, levels = levels(test$Class))
confusionMatrix(xgb_pred, test$Class, positive = "malignant")
```

<br>

Consistent with previous models, `Cell.shape`, `Cell.size`, and `Bare.nuclei` ranked among the most important features. The consistency of variable importance across models suggests that these predictors carry the most discriminative information for breast cancer classification.

The model achieved an **accuracy of 97.1%** and a **sensitivity of 94.4%**, slightly lower than logistic regression and Random Forest, misclassifying four malignant cases as benign.



<br><br><br><br>

### 6. Model Evaluation

<br><br>

Model performance was evaluated across all five classifiers using confusion matrices and ROC/AUC analysis. <br
Given the clinical context, **sensitivity** was treated as the primary evaluation metric, as failing to detect a malignant tumor carries significantly greater clinical risk than a false positive.

<br><br>

#### 6-1. Confusion Matrix

<br>

The table below summarizes key performance metrics for all five models on the test set.

<br>

```{r confusion_summary, echo=FALSE}
results <- data.frame(
  Model       = c("Logistic Regression", "LASSO", "Decision Tree", 
                  "Random Forest", "XGBoost"),
  Accuracy    = c(0.9755, 0.9706, 0.9412, 0.9804, 0.9706),
  Sensitivity = c(0.9577, 0.9437, 0.9155, 0.9718, 0.9437),
  Specificity = c(0.9850, 0.9850, 0.9549, 0.9850, 0.9850),
  FN          = c(3, 4, 6, 2, 4)
)

knitr::kable(results, 
             caption = "Model Performance Comparison on Test Set",
             align   = c("l", "c", "c", "c", "c"))
```

<br>

Random Forest achieved the highest sensitivity (97.2%), misclassifying only two malignant cases as benign. Logistic Regression followed with a sensitivity of 95.8%.

LASSO and XGBoost both achieved a sensitivity of 94.4%, slightly lower than that of Logistic Regression and Random Forest. The Decision Tree exhibited the lowest sensitivity (91.6%), reflecting the limitation of relying on only two predictors.

All models substantially outperformed the No Information Rate (65.2%), indicating that each model provides meaningful discriminative power beyond naive classification.

<br><br><br>


#### 6-2. ROC / AUC Comparison

<br>

ROC curves were generated for all five models to evaluate discriminative performance across all possible classification thresholds. 
The Area Under the Curve (AUC) provides a threshold-independent summary of each model's ability to distinguish malignant from benign cases.

<br>

```{r roc_auc, fig.width=8, fig.height=6}
roc_logit <- roc(test$Class, logit_prob,
                 levels = c("benign", "malignant"))
roc_lasso <- roc(test$Class, as.numeric(lasso_prob),
                 levels = c("benign", "malignant"))
roc_tree  <- roc(test$Class, predict(tree_pruned, newdata = test, type = "prob")[, "malignant"],
                 levels = c("benign", "malignant"))
roc_rf    <- roc(test$Class, rf_prob,
                 levels = c("benign", "malignant"))
roc_xgb   <- roc(test$Class, xgb_prob,
                 levels = c("benign", "malignant"))

auc_df <- data.frame(
  Model = c("Logistic", "LASSO", "Decision Tree", "Random Forest", "XGBoost"),
  AUC   = round(c(auc(roc_logit), auc(roc_lasso), auc(roc_tree),
                  auc(roc_rf),    auc(roc_xgb)), 4)
)
print(auc_df)

plot(roc_logit, col = "blue",   lwd = 2, main = "ROC Curve Comparison")
plot(roc_lasso, col = "red",    lwd = 2, add = TRUE)
plot(roc_tree,  col = "green",  lwd = 2, add = TRUE)
plot(roc_rf,    col = "purple", lwd = 2, add = TRUE)
plot(roc_xgb,   col = "orange", lwd = 2, add = TRUE)

legend("bottomright",
       legend = paste(auc_df$Model, "=", auc_df$AUC),
       col    = c("blue", "red", "green", "purple", "orange"),
       lwd    = 2,
       cex    = 0.7,
       bty    = "n")
```

<br>

Logistic Regression achieved the highest AUC (0.9988), followed closely by LASSO (0.9987) and Random Forest (0.9985). XGBoost also performed strongly with an AUC of 0.9978. The Decision Tree showed a notably lower AUC (0.9538), consistent with its reliance on only two predictors.

The marginal AUC differences among the top four models suggest that the strong discriminative signal in this dataset is primarily driven by the underlying cytological features, rather than by differences in modeling approach.

When considered alongside sensitivity, Random Forest can be regarded as the best overall model, achieving the highest sensitivity (97.2%) while maintaining a near-optimal AUC (0.9985).

<br><br><br><br>



### 7. Conclusion

<br>

This project applied and compared five classification models, including Logistic Regression, LASSO, Decision Tree, Random Forest, and XGBoost, for breast cancer diagnosis using the UCI Breast Cancer Wisconsin dataset.

<br>

Across all models, `Cell.shape`, `Cell.size`, and `Bare.nuclei` consistently emerged as 
the most discriminative predictors, a finding supported by EDA, variable importance plots, 
and the decision tree structure. `Mitoses`, in contrast, showed limited predictive value 
and was eliminated by LASSO through coefficient shrinkage.

<br>

Among the five models, **Random Forest** can be considered the best overall performer, achieving the highest sensitivity (97.2%) and accuracy (98.0%), misclassifying only two malignant cases as benign. Although Logistic Regression achieved a marginally higher AUC (0.9988), the difference was negligible in practice, and its superior sensitivity makes Random Forest more suitable in this clinical context.

<br>

The Decision Tree, despite its lower predictive performance, offered the most interpretable output in the form of a simple two-variable rule that could be directly translated into clinical screening guidelines.

<br>

Several limitations should be acknowledged. The dataset is relatively small (n = 683) and 
based on cytological assessments from a single institution, which may limit generalizability. 
Additionally, the use of a single train/test split without repeated cross-validation means 
that performance estimates may be sensitive to the specific data partition. Future work 
could explore repeated cross-validation, threshold optimization to further improve 
sensitivity, and validation on independent external datasets.


<br><br><br>

